---
title: "Langford Sequences"
description: |
  Describing Langford and Skolem sequences and how to search for them in R.
author:
  - name: James Curley
    url: jamescurley.blog
categories:
  - puzzles
date: 10-16-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

In 1958, the mathematician C.D. Langford was watching his son play with blocks. He wrote:

> There were two of each color, and one day I noticed that he had placed them in a single pile so that between the red pair there was one block, two between the blue pair, and three between the yellow. I then found that by a complete rearrangement I could add a green pair with four between them.

<br>

```{r}
library(png)
library(grid)
img <- readPNG("../../img/langford2.png")
grid.raster(img)
```

<br>

The general problem was summarized by Hayasaka & Saito (1979) as: *“Given $2n$ numbers, two each of the numbers $1, 2, …., n$ to find whether they can be arranged in a row in such a way that the two occurrences of each number $k$ are separated by exactly $k$ other elements of the row*.

Langford actually discovered several examples for various different number of blocks. Here are some examples of his sequences:

```{r}
img <- readPNG("../../img/langford3.png")
grid.raster(img)
```

<br>

What one might notice is that all the solutions are for a number of blocks $n$ that are divisible by 4, or one less than a number that is divisible by 4. A bit more explanation for why this is so can be found at [Nick Berry’s excellent blog](http://datagenetics.com/blog/october32014/index.html).


As $n$ increases, there are an increasing number of Langford Sequences. There is only one solution for $n=3$ & $n=4$, 26 solutions for $n=7$, 150 for $n=8$, 17,792 for $n=11$, 108,144 for $n=12$ ….. 46,845,158,056,515,936 for $n=24$ ! Obviously, trying to find all solutions for any given $n$ is computationally intensive!

Although Langford sequences can only be found when $n$ is divisible by 4 or when $n+1$ is divisible by 4, there is an alternative type of Langford sequence called a Hooked Langford Sequence which can be found for all values of $n$. A Hooked Langford Sequence is one where another block/letter is inserted in the penultimate position in the sequence.

At the same time as Langford wrote his piece, a related phenomenon was being independently researched by a Swedish mathematician, Thoralf Skolem, who was trying to form sequences of $2n$ blocks with intervals of $0,1,2,3 ... n$.

My aim here is to write a small script that can find these sequences.  Below is an image of each type of these sequences:

<br>

```{r, preview=TRUE}
img <- readPNG("../../img/langford.png")
grid.raster(img)
```

<br>


### Langford's Sequences

Say we have $2*8$ blocks, let’s represent them with letters. We can then generate a randomly sample sequence of these letters as a starting point. A random shuffle is unlikely to be a Langford sequence. We an check by finding the differences between the positions/indexes of each letter in the vector:


<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
x <- rep(LETTERS[1:8],2)
set.seed(1)
y <- sample(x)
y
```
<br>

This gets the difference (number of letters between) the pairs of A's:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
abs(diff(which(y=="A")))-1  #6
```
<br>

Likewise for the D's and the H's:


<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
abs(diff(which(y=="D")))-1  #10
abs(diff(which(y=="H")))-1  #1
```
<br>

We can get the results for all the individual letters with a loop:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
res<-NULL
for(i in 1:8){  res[[i]] <- abs(diff(which(y==LETTERS[i])))-1 }
          
names(res)<-LETTERS[1:8]
res
```
<br>


This is fine for one randomly sampled sequence, but when finding Langford sequences, we might want to check many sequences, so it is better to have a faster function. `match()` seems to be pretty fast for this purpose:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
langford <- function(x) {
  ux = unique(x)
  mux = match(x, ux)      
  ans = integer(length(ux))       
  for(i in seq_along(x)) ans[mux[i]] = i - ans[mux[i]]        
  return(setNames(ans - 1L, ux))
}

langford(y)
```
<br>


### Finding a Langford Sequence

Say we wanted to find a solution for 8 blocks. My strategy here is to keep sampling sequences until one satisfies the criteria. The way to check whether a sequence is a Langford sequence is to sort the intervals between each pair of letters and then ask if this is equal to `1:8`. If it is, then we break and return that sequence. 

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
set.seed(17)
while(TRUE) {
  
  samps <- sample(x)
  tmp <- sort(langford(samps))
  
  if(sum(tmp==c(1:8))==8)
    
    break
}

samps
```
<br>

We can see that this is a Langford Sequence:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
sort(langford(samps))  ## the interval between pairs of letters in the sequence```
```
<br>

This is obviously a brute force approach, and not a cute algorithm that optimally finds solutions. We can generalize for any ‘n’:


<br>
  
  
```{r, warning=FALSE, message=FALSE, echo=TRUE}
get.langford <- function(n){

  
  
while(TRUE) {

  x <- rep(LETTERS[1:n],2)
  samps <- sample(x)
  tmp <- sort(langford(samps))
  
  if(sum(tmp==c(1:n))==n)
    
    break
}

return(samps)
}

```
<br>

I could add some `if-stop` ‘s to this to prevent someone from entering a ’n’ that will not work (e.g. 5,6). Also, if someone wanted to use an ‘n’ of >26 then we’d have to use combinations of letters. I don’t recommend this though, as the function is pretty slow for anything greater than n=12.

I think the strategy could also be optimized such that rather than randomly sampling sequences, the sequences tested are done so in a more systematic fashion.

Nevertheless, here it is working for n=12.

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
set.seed(77)
lang12 <- get.langford(12)
lang12
```
<br>


<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
sort(langford(lang12))
```
<br>

We can try and plot this with ggplot:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
library(tidyverse)

# dataframe of position of each letter           
df <- data.frame(pos = 1:24, let = lang12)


# get distances between each letter
lang12dif <- sort(langford(lang12))

# record start and end position of each letter
# to make line between them
df1 <- df %>% group_by(let) %>% filter(pos==min(pos))%>%
  full_join(data.frame(
           let = names(lang12dif),
           diff = lang12dif
           )
            ) %>%
  mutate(end = diff+pos+1)
           

# reorder levels so can arrange in order
df$let <- factor(df$let, levels=names(lang12dif))
df1$let <- factor(df1$let, levels=names(lang12dif))

df %>%
  ggplot() +
  geom_segment(data=df1, aes(x=pos, xend=end, y=let, yend=let), color="gray44")+
  geom_tile(data=df, aes(x=pos, y=let, fill=let), color='black') +
  geom_text(data=df, aes(x=pos, y=let, label=let)) +
  theme_classic() +
  geom_text(data=df1, aes(x=(pos+end)/2, y=let, label=diff),color='red', nudge_y = 0.25) +
  theme(legend.position = 'none')
  

```
<br>


<br>
<br>

### Finding Skolem Sequences

This is simply just a change in the if statement prior to the break. Here we are looking for sequences that start with a gap of 0, and not 1.

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
skolem <- function(n){

while(TRUE) {

  x <- rep(LETTERS[1:n],2)
  samps <- sample(x)
  tmp <- sort(langford(samps))
  
  if(sum(tmp==c(0:(n-1)))==n)
    
    break
}

return(samps)
}
```
<br>

For instance, a Skolem sequence for n=8:
<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
set.seed(88)
sk <- skolem(8)
sk

sort(langford(sk))

```
<br>
<br>

### Finding Hooked Langford Sequences

AA Hooked Langford Sequence is one where another block/letter is inserted in the penultimate position in the sequence. They can be found for any $n$:

The code for these Hooked Langford Sequence is here:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
hooked.langford <- function(n){

while(TRUE) {

  x <- rep(LETTERS[1:n],2)
  samps <- sample(x)
  samps <- c(head(samps,((n*2)-1)), "X", tail(samps, 1))

  
  tmp <- sort(langford(samps))
  
  if(sum(tmp==c(1:n))==n)
    
    break
}

return(samps)
}
```
<br>
Some examples:

<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}
set.seed(1)
h1 <- hooked.langford(5)
h1

sort(langford(h1))


```
<br>


<br>
```{r, warning=FALSE, message=FALSE, echo=TRUE}

set.seed(11)
h2 <- hooked.langford(9)
h2

sort(langford(h2))
```
<br>


