---
title: "TSA passenger screenings during Covid"
description: |
  Examining how total USA airline passenger changes following the Covid pandemic
author:
  - name: James Curley
    url: jamescurley.blog
categories:
  - time_series
date: 10-20-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Clearly air traffic plummeted during Covid. [Bob Rudis tweeted](https://twitter.com/hrbrmstr/status/1318470131270418433) a great figure showing how passenger numbers changed between 2019 and 2020. I wanted to recreate the figure here. It's a good exercise in web scraping data, data cleaning, and how to plan and actually make appealing and informative time-series line graphs with customized `ggplot2` elements.


### The data

The data are available at the TSA [website here](https://www.tsa.gov/coronavirus/passenger-throughput). They appear to be updated daily, so hopefully the code below should work no matter what day the data are downloaded.

We will use the `rvest` package to read in the url and scrape the HTML table. The `extract2()` function comes from `magrittr` and helps us get the first (and only) element from the list produced by `html_table()`.

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(rvest)
library(tidyverse)
library(magrittr)

url <- "https://www.tsa.gov/coronavirus/passenger-throughput"

df <- read_html(url) %>%
  html_table() %>%
  extract2(1)

head(df)

```
<br>

We have three columns: date, total passengers in 2020, and total passengers on the corresponding weekday of 2019.  So, the final column is not the exact same calendar data - but the corresponding closest Monday, Tuesday, etc. of 2019.

First, let's clean up the column names:

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE}
colnames(df) <- c("date", "pass2020", "pass2019")
head(df)
```
<br>

We still have some cleaning steps. We need to fix the date, the numbers and to reshape the dataframe.

Let's make the date a date object:

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE}
df$date <- as.Date(df$date, format="%m/%d/%Y")
str(df) #it's now a 'date' column
```
<br>
Next, we'll remove commas from the numbers and ensure the columns are numeric - you can see from the output above that R still thinks they are characters:

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE}

df$pass2020 <- as.numeric(gsub(",", "", df$pass2020))
df$pass2019 <- as.numeric(gsub(",", "", df$pass2019))
head(df)
```
<br>

To be able to plot these data with `ggplot2`, we need to convert this dataframe from a wide dataframe to a long dataframe:


<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE}

df <- df %>% 
  pivot_longer(cols=2:3)

df
```
<br>

Now we can graph.  We'll first make the basic graph, and then we can add the customization:

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE, fig.width=8, fig.height=5}
ggplot(df, aes(x=date, y=value, color=name)) + 
  geom_line()
```
<br>

There are lots of interesting patterns in that graph.  Clearly, number one is that the passenger numbers in mid March 2020 showed a dramatic fall, and a very slow recovery. The rate of recovery seems to have slowed down between July 2020 and October 2020. Starting in October 2020 we're perhaps seeing a slightly increase again in the rate of passenger increase.  The other patterns of interest are the weekly cycle of passenger numbers - an obvious pattern, but always notable to observe, as well as the season variation in traffic that usually occur in normal years.

To make this chart more readable, we want to change the passenger numbers on the y-axis from 'e' numbers to understandable numbers, add a title, remove the legend, change the theme and the line colors.  I will also make the lines thicker and change the font size of the text on the figure. We should also source the data and original author.


<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE, fig.width=8, fig.height=5}
library(scales) # needed to reformat the y-axis with commas in the numbers

ggplot(df, aes(x=date, y=value, color=name)) + 
  geom_line(lwd=1) +
  scale_color_manual(values=c("#123abc", "#f59a46"))+
  scale_y_continuous(breaks = seq(0,3000000, 500000), labels = comma) +
  ylab("") +
  xlab("") +
  ggtitle("TSA checkpoint travel numbers for 2020 and 2019",
          subtitle = "Total Traveler Throughput; Same Weekday") +
  theme_minimal() +
  labs(caption = "Adapated from Figure by Bob Rudis; Data: https://www.tsa.gov/coronavirus/passenger-throughput") +
  theme(
    legend.position = 'none',
    axis.text = element_text(size=12),
    plot.title = element_text(size=18)
    )
```
<br>

This looks a lot better but it still has one major issue, which is you don't know which line is 2019 and which line is 2020. Well, you could take a guess, but it should be keyed in the figure.  I want to change colors in the title of the plot to indicate this.

This is a little trickier and requires a package called `ggtext` which is downloadable [from here](https://github.com/wilkelab/ggtext).  The code you need to include requires you to use a little bit of HTML tagging, but the overall effect is worth it.  More examples of using `ggtext` can be found on it's GitHub page.  [Here is a useful example](https://stackoverflow.com/questions/49735290/ggplot2-color-individual-words-in-title-to-match-colors-of-groups) also on stackoverflow.

<br>
```{r, message=FALSE, warning=FALSE, echo=TRUE, fig.width=8, fig.height=5, preview = TRUE}
library(ggtext) #remotes::install_github("wilkelab/ggtext")

ggplot(df, aes(x=date, y=value, color=name)) +
  geom_line(lwd = 1) +
  scale_color_manual(
    name = NULL,
    values = c(pass2019 = "#f59a46", pass2020 = "#123abc"),
    labels = c(
      pass2019 = "<i style='color:#f59a46'>I. pass2019</i>",
      pass2020 = "<i style='color:#123abc'>I. pass2020</i>")
  ) +
  labs(
    title = "**TSA checkpoint travel numbers for <span style='color:#123abc;'>2020</span> and <span style='color:#f59a46;'>2019</span>**  
    <span style='font-size:11pt'>Total Traveler Throughput; Same Weekday</span>
    ",
    x = "Sepal length (cm)", y = "Sepal width (cm)"
  ) +
  scale_y_continuous(breaks = seq(0,3000000, 500000), labels = comma) +
  ylab("") +
  xlab("")+
  theme_minimal() +
  labs(caption = "Adapated from Figure by Bob Rudis; Data: https://www.tsa.gov/coronavirus/passenger-throughput")+
  theme(
    plot.title = element_markdown(lineheight = 1.1),
     legend.position = 'none'
  )

```
<br>


<br>
