---
title: "World Cup Birthday Problem"
description: |
  Do world cup squad members share birthdays?
author:
  - name: James Curley
    url: jamescurley.blog
categories:
  - puzzles
  - regex
  - soccer
date: 10-14-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<br><br>

### The Birthday Problem

The Birthday Problem is quite a well known puzzle. If you attended a party of 20 people or so and were asked how likely it was that any two people at the party shared a birthday - what would you think?  Most people when they are first confronted with this question think the probability will be quite small.  After all, there are 365 days in the year, and only 20 people at the party.  However, it turns out that the chances are pretty decent.  

There is, obviously, some probability mathematics behind this that will tell us that the chances of any two people sharing a birthday at the party of 20 people is around 41%.  You can read more about [the math here](https://en.wikipedia.org/wiki/Birthday_problem).  

Instead of going through the math, I'd like to show this visually using a simulation.  Then we'll check this problem against some real world data - using soccer world cup squads.

<br>

#### Simulating the Birthday Problem

What we're doing below is to use `sample()` to randomly select `n` number of numbers between 1 and 365. These are our days of the year - we're just going to ignore leap years. Importantly, we're selecting these numbers WITH replacement. 

For example, let's select five numbers between 1 and 365 with replacement

```{r}
set.seed(1)
```

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}
sample(1:365, 5, T)
```
<br>

As you can see from these numbers, none are duplicated.  However, if we got 25 numbers, then we have more chance of getting some numbers twice:


```{r message=FALSE, warning=FALSE, echo=TRUE}
vs <- sample(1:365, 25, T)
vs
```
<br>

We next use the function `duplicated()` to find out if any of the numbers are .... duplicated.

```{r message=FALSE, warning=FALSE, echo=TRUE}
duplicated(vs)
```
Here, the 21st number is duplicated - which is 141. It appears twice in the list - this would correspond to a shared birthday of May 20th (the 141st day of the year).

<br>

The code below runs 10,000 lots of group sizes of 1-50. For each group size, and for each permutation, we count how many of the birthdays are duplicated:

<br>

```{r}
set.seed(1)
```


```{r message=FALSE, warning=FALSE, echo=TRUE}

nperm <- 20000
maxn <- 50

res <- matrix(NA, nrow = maxn, ncol = nperm)

for(n in 1:maxn) {
  for(i in 1:nperm) {
    x <- sample(1:365, n, T)
    res[[n,i]] <- sum(duplicated(x))
  }
}

```
<br>

The table below visualizes the first 5 iterations of group sizes of 1-15.  Obviously, if it's a party of one person it's a lame party and also nobody is going to share a birthday. What you can see in the results below, is that in the 3rd run, people shared birthdays when there were 8 or 15 people at the party. In the 4th run, people shared birthdays when there were 12 or 14 people at the party.

<br>

```{r message=FALSE, warning=FALSE, echo=TRUE}
# first 5 iterations of group sizes 1-15
res[1:15,1:5]
```

<br>

For each group size, we can count over all 20,000 iterations how many of the 20,000 instances had at least two people that shared a birthday.  We can then work out the proportion of times this happens:

<br>


```{r message=FALSE, warning=FALSE, echo=TRUE}
v <- apply(res, 1, function(x) sum(x>0))
round(v/nperm,2)
```

<br>

This shows you that when there are only 5 people at the party, you have a 3% chance of there being shared birthdays.  When there are 40 people at a party, there is a 89% chance of there being a shared birthday.  It turns out that when there are 23 people at a party there is a 51% chance of there being a shared birthday. 23 is the number of people where it's more likely for there to be people that share a birthday than not.


We can make a simple visualization of this. The solid black line represents the probability (y-axis) of at least two people sharing a party for a given group size (x-axis):

<br>



```{r message=FALSE, warning=FALSE, echo=TRUE}
plot(1:maxn, v/nperm, type='l', lwd=2,
     xlab="Party Size", ylab="Probability of Birthday Being Shared")
abline(h=0.5, col="red",lty=2)
abline(v=23, col="black",lty=3)
```
<br>
<br>


#### World Cup Data


When soccer teams compete at major championships, coaches have to pick squads of 23 for the tournament. Obviously, based on the above, this is a fairly interesting number.  It led me to wonder, what proportion of world cup squads have players that shared birthdays?  

The 32 squads for the 2018 world cup are [available here](https://en.wikipedia.org/wiki/2018_FIFA_World_Cup_squads).  What I'm doing below is to use the R package `rvest` to scrape these tables from wikipedia. 

<br>

```{r message=FALSE, warning=FALSE, echo=TRUE}
library(rvest)    
URL <- "https://en.wikipedia.org/wiki/2018_FIFA_World_Cup_squads"

squads <- URL %>% 
  read_html %>%
  html_nodes("table") %>%
  html_table() %>%
  head(32)


```

<br>

The squads are stored in a list. Let's look at the first four rows of the seventh squad, which is Portugal.  It's a bit ugly to look at because it's split over several rows, but hopefully you can see what type of data we have:  


```{r message=FALSE, warning=FALSE, echo=TRUE}
squads[[7]][1:4,]

```
<br>

Each table has the following column names:

```{r message=FALSE, warning=FALSE, echo=TRUE}
colnames(squads[[1]])
```

<br>

As you might notice, the date of birth column is a bit of a mess, which we'll deal with shortly.  There is also no 'country' column, which would be useful.  To create this, I realized that on the same wikipedia page, if we grabbed all text that was written with a `<h3>` HTML tag, we could get the country names and add them in to each squad - after getting rid of a bit of extraneous text that comes along at the end of each them when we scrape the data.

Shown below are the first 4 rows of the 23rd squad, South Korea, once we've added in the country name: 

<br>

```{r message=FALSE, warning=FALSE, echo=TRUE}

names(squads) <- URL %>% 
  read_html %>%
  html_nodes("h3") %>%
  html_text() %>%
  head(32) %>%
  gsub("\\[edit\\]","",.)

squads <- Map(cbind, squads, country = names(squads))

squads[[23]][1:4,]
```
<br>

What we really care about for this post is whether players in the same squad share a birthday. Let's illustrate this looking at South Korea's birthdays.

Here are all the birthdays as listed in the date of birth column:


<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}
x <- squads[[23]]$`Date of birth (age)`
x
```
<br>

We don't need all this stuff, so we can grab everything inside the brackets using a regex. To be honest, I'm sure there is an easier way of doing this - this is just the way I could figure out quickly:

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}

#get all inside brackets
gsub("\\(([^()]*)\\)|.", "\\1", x, perl=T)

```
<br>

Clearly, this is still too much info.  However, because the data is so uniform in type, we can use `substr()` to get the text from the 6th to 10th character of each string. That corresponds to the month and day of birth (years aren't important for this problem):

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}

#day-month
vec <- substr(gsub("\\(([^()]*)\\)|.", "\\1", x, perl=T),6,10)
vec

```

<br>

We can used `duplicated()` to now return all the dates that are duplicates:

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}

duplicated(vec) | duplicated(vec, fromLast=TRUE)
vec[duplicated(vec) | duplicated(vec, fromLast=TRUE)]

```
<br>

So, in the South Korean squad, there are two players who share birthdays on 27th February. These players happen to be Koo Ja-cheol & Kim Young-gwon - but this is all revealed at the bottom.

<br>

The below is a custom function using the above logic to check each squad for duplicated birthdays and to put them into each squad's dataframe:

```{r message=FALSE, warning=FALSE, echo=TRUE}

dob_dup <- function(df){
  x <- df$`Date of birth (age)`
  vec <- substr(gsub("\\(([^()]*)\\)|.", "\\1", x, perl=T),6,10)
  df$daymonth <- vec
  df$dupl <- duplicated(vec) | duplicated(vec, fromLast=TRUE)
  return(df)
}

```

<br>

Here, we apply the function to all squads (using `map()` from `tidyverse` to be able to apply the function over each squad dataframe in the list). I'm showing you Morocco's first five players as an example:

<br>

```{r message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)
squads <- squads %>% map(dob_dup)
squads[[6]][1:5,]
```
<br>

Applying this to every squad, we can then count up the number of players that share a birthday in each squad of 23 players:

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}

v <- squads %>% map(~ sum(.$dupl)) %>% unlist()
v
```
<br>

Poland lead the way with 8 players sharing birthdays !  Here they are:


<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}
squads[[31]] %>% filter(dupl==T) %>% select(1:3,5,7,9)
```
<br>


So, how many of the thirty-two 23 man squads had at least two players that shared a birthday ?

<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}
v[v>0]
```
<br>

It turns out that 16/32 squads had players that shared birthdays - so 50% which is basically what our simulation above told us was the probability.

However, this is not probably the full story.  You may notice that we have more shared birthdays than two in many squads. Interestingly, it turns out that soccer squads may not be completely random samples with respect to birthdays. Often in sports teams, soccer included, players tend to be born [in certain months of the year](https://en.wikipedia.org/wiki/Relative_age_effect). This is often due to older children in age groups being picked for school teams and so on.  We may well have some sampling bias in our data - something we can explore in another post.

<br>

Out of interest, here is the full list of all players that share birthdays in squads:


<br>
```{r message=FALSE, warning=FALSE, echo=TRUE}

squads %>% map(dob_dup) %>% map(~filter(., dupl==T)) %>%
  map(~ select(., c(3, 5, 7:9))) %>% data.table::rbindlist()

```
<br>
<br>
<br>

